Procrastination as a Marker of Cognitive Decline: Evidence from Longitudinal Transitions in the Older Adult Population

### 2025-10-03 manuscript/manuscript.qmd
Update :fire:

---
title: "Procrastination as a Marker of Cognitive Decline: Evidence from Longitudinal Transitions in the Older Adult Population"
author: 
  - name: Cormac Monaghan
    url: https://c-monaghan.github.io/
    email: cormacmonaghan@protonmail.com
    orcid: https://orcid.org/0000-0001-9012-3060
    affiliation:
      - Hamilton Institute, Maynooth University, Ireland
      - Department of Psychology, Maynooth University, Ireland
    attributes:
      corresponding: true
  - name: Rafael de Andrade Moral
    url: https://rafamoral.github.io/
    orcid: https://orcid.org/0000-0002-0875-3563
    affiliation: Department of Mathematics and Statistics, Maynooth University, Ireland
  - name: Michelle Kelly
    url: https://www.ncirl.ie/About/A-Z-Staff-Directory/Staff/346
    orcid: https://orcid.org/0000-0001-6159-6707
    affiliation: Department of Psychology, National College of Ireland
  - name: Joanna McHugh Power
    url: https://www.maynoothuniversity.ie/people/joanna-mchugh-power
    orcid: https://orcid.org/0000-0002-7387-3107
    affiliation: Department of Psychology, Maynooth University, Ireland
correspondence-prefix: "Correspondence concerning this article should be addressed to"
abstract: |
  Cognitive decline is a global health concern, making the identification of early, modifiable risk factors essential. While apathy is a recognized prodromal marker, procrastination may also signal early executive dysfunction. We used longitudinal secondary data from the United States Health and Retirement Study among adults aged 60+ $(n = 549; \bar{x} = 69.70;\sigma = 7.58)$. Cognitive function, procrastination, and a proxy measure of apathy were assessed. Transitions between normative cognitive function, mild cognitive impairment (MCI), and dementia were modeled using a discrete-time first-order Markov model. Procrastination scores were higher among individuals with MCI or dementia than those with normative cognitive function. Procrastination also interacted with age, disproportionately increasing the risk of decline in the oldest participants. Procrastination was associated with cognitive impairment and predicted transitions to MCI, suggesting it may serve as both an early behavioral marker and compounding risk factor.
additional-info: |
  **Data Availability:** All data and replication code are publicly available at <https://github.com/C-Monaghan/crisp-clam>
keywords:
  - Dementia risk
  - Mild cognitive impairment
  - Procrastination
  - Apathy
  - Executive dysfunction
  - Aging;
code-repo: "Access the code, data, and analysis at <https://github.com/C-Monaghan/crisp-clam>"
bibliography: references.bib
csl: csl/apa.csl
link-citations: true
editor: 
  markdown: 
    wrap: 80
---

@@

library(ggplot2)

@@

Dementia is characterized by progressive decline of cognitive function, leading
to memory loss and difficulties with daily living [@prince2013;
@sanz-blasco2022]. The global burden of dementia is substantial, with cases
projected to rise to 152.8 million by 2050 [@nichols2022]. Identifying and
addressing modifiable risk factors is crucial to mitigate this projected rise in
prevalence. Two particularly vulnerable groups are older adults and those with
mild cognitive impairment (MCI), of whom up to 80% progress to dementia within
six years \[@cooper2015; @shigemizu2020 @tschanz2006}.

@@

@@

```{r}
#| label: tbl-descriptives
#| fig-width: 12
#| tbl-cap: "Baseline descriptives for the study sample and stratified by cognitive status category."
#| tbl-cap-location: top
#| include: true

# Table information
tbl_descriptives <- tibble::tribble(
  ~"", ~"Full sample <br>(n = 549)", ~"NC <br>(n = 452)", ~"MCI <br>(n = 86)", ~"Dementia <br>(n = 11)",
  "Age (years)", "69.70 ± 7.58", "69.70 ± 7.55",	"69.60 ± 7.55",	"70.00 ± 9.89",
  "**Sex**", " ", " ", " ", " ",
  "Male", "38.43% <br>(n = 211)", "39.82% <br>(n =180)", "32.56% <br>(n = 28)", "27.27% <br>(n = 3)",
  "Female", "61.57% <br>(n = 338)", "60.18%<br>(n = 272)", "67.44% <br>(n = 58)",	"72.73 <br>(n = 8)",
  "**Education**", " ", " ", " ", " ",
  "No degree",	"16.24% <br>(n = 98)", "10.56% <br>(n = 47)",	"39.53% <br>(n = 34)",	"63.54% <br>(n = 7)",
  "GED", "51.58% <br>(n = 279)",	"51.91% <br>(n = 231)",	"52.33% <br>(n = 45)",	"27.27 <br>(n = 3)",
  "Further <br>education", "32.29% <br>(n = 175)", "37.53% <br>(n = 167)", "8.14% <br>(n = 7)",	"9.09% <br>(n = 1)",
  "Apathy",	"0.37 ± 0.63", "0.31 ± 0.59",	"0.65 ± 0.76",	"0.36 ± 0.51",
  "Procrastination",	"28.60 ± 12.00", "27.70 ± 11.30",	"32.10 ± 13.80",	"39.00 ± 16.00"
)

@@

# Outputting table 
tbl_descriptives |>
  setNames(c(" ", "Full sample <br>(n = 549)", "NC <br>(n = 452)", "MCI <br>(n = 86)", "Dementia <br>(n = 11)")) |>
  tinytable::tt(notes = list(
    caption, 
    a = list(i = 10, j = 1, text = footnote))
    ) |>
  tinytable::style_tt(j = 2:5, align = "c") |>
  tinytable::format_tt("notes", markdown = TRUE) |>
  tinytable::format_tt(j = 1, markdown = TRUE)
```

```{r}
#| label: fig-river
#| fig-width: 15
#| fig-height: 10
#| fig-cap: "Frequency of cognitive status transitions between HRS waves."
#| include: true

@@

```{r}
#| label: fig-markov
#| fig-width: 12
#| fig-height: 10
#| fig-cap: "Odds ratios (95% Confidence Intervals) of transitioning from one cognitive state to another calculated using generalised logit models. <br>*Note*. Odds ratios significantly different from 1 at a 5% significance rate are presented in blue (positive) or orange (negative),  otherwise they are presented in gray. NC = Normative cognitive function;  MCI = Mild cognitive impairment."
#| include: true

@@

```{r}
#| label: fig-age-preds
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Predicted transition probabilities by procrastination and age. <br>*Note.* Each curve represents a different age profile, ranging from the minimum (62 years) to the maximum (97 years) of the dataset. Specific ages of interest (70, 80, and 90) are highlighted."
#| include: true

>>>>

---
title: "Procrastination as a Marker of Cognitive Decline: Evidence from Longitudinal Transitions in the Older Adult Population"
author: 
  - name: Cormac Monaghan
    url: https://c-monaghan.github.io/
    email: cormacmonaghan@protonmail.com
    orcid: 0000-0001-9012-3060
    affiliation: Maynooth University
    attributes:
      corresponding: true
  - name: Rafael de Andrade Moral
    url: https://rafamoral.github.io/
    orcid: 0000-0002-0875-3563
    affiliation: Maynooth University
  - name: Michelle Kelly
    url: https://www.ncirl.ie/About/A-Z-Staff-Directory/Staff/346
    orcid: 0000-0001-6159-6707
    affiliation: National College of Ireland
  - name: Joanna McHugh Power
    url: https://www.maynoothuniversity.ie/people/joanna-mchugh-power
    orcid: 0000-0002-7387-3107
    affiliation: Maynooth University
correspondence-prefix: "Correspondence concerning this article should be addressed to"
abstract: |
  Cognitive decline is a global health concern, making the identification of early, modifiable risk factors essential. While apathy is a recognized prodromal marker, procrastination may also signal early executive dysfunction. We used longitudinal secondary data from the United States Health and Retirement Study among adults aged 60+ $(n = 549; \bar{x} = 69.70;\sigma = 7.58)$. Cognitive function, procrastination, and a proxy measure of apathy were assessed. Transitions between normative cognitive function, mild cognitive impairment (MCI), and dementia were modeled using a discrete-time first-order Markov model. Procrastination scores were higher among individuals with MCI or dementia than those with normative cognitive function. Procrastination also interacted with age, disproportionately increasing the risk of decline in the oldest participants. Procrastination was associated with cognitive impairment and predicted transitions to MCI, suggesting it may serve as both an early behavioral marker and compounding risk factor.
additional-info: |
  **Data Availability:** All data and replication code are publicly available at <https://github.com/C-Monaghan/crisp-clam>
keywords:
  - Dementia risk
  - Mild cognitive impairment
  - Procrastination
  - Apathy
  - Executive dysfunction
  - Aging;
code-repo: "Access the code, data, and analysis at <https://github.com/C-Monaghan/crisp-clam>"
bibliography: references.bib
csl: csl/apa.csl
link-citations: true
editor: 
  markdown: 
    wrap: 80
keep-tex: true
---

@@

library(ggplot2)
library(tinytable)

@@

Dementia is characterized by progressive decline of cognitive function, leading
to memory loss and difficulties with daily living [@prince2013;
@sanz-blasco2022]. The global burden of dementia is substantial, with cases
projected to rise to 152.8 million by 2050 [@nichols2022]. Identifying and
addressing modifiable risk factors is crucial to mitigate this projected rise in
prevalence. Two particularly vulnerable groups are older adults and those with
mild cognitive impairment (MCI), of whom up to 80% progress to dementia within
six years [@cooper2015; @shigemizu2020 @tschanz2006].

@@

\clearpage

@@

```{r}
#| label: tbl-descriptives
#| fig-width: 12
#| tbl-cap: "Baseline descriptives for the study sample and stratified by cognitive status category."
#| tbl-cap-location: top
#| tbl-pos: H
#| include: true

# Table information
tbl_descriptives <- tibble::tribble(
  ~"", ~Full, ~NC, ~MCI, ~Dementia,
  "Age (years)", "69.70 ± 7.58", "69.70 ± 7.55",	"69.60 ± 7.55",	"70.00 ± 9.89",
  "Male", "38.43% (n = 211)", "39.82% (n =180)", "32.56% (n = 28)", "27.27% (n = 3)",
  "Female", "61.57% (n = 338)", "60.18% (n = 272)", "67.44% (n = 58)",	"72.73% (n = 8)",
  "No degree",	"16.24% (n = 98)", "10.56% (n = 47)",	"39.53% (n = 34)",	"63.54% (n = 7)",
  "GED", "51.58% (n = 279)",	"51.91% (n = 231)",	"52.33% (n = 45)",	"27.27% (n = 3)",
  "Further education", "32.29% (n = 175)", "37.53% (n = 167)", "8.14% (n = 7)",	"9.09% (n = 1)",
  "Apathy",	"0.37 ± 0.63", "0.31 ± 0.59",	"0.65 ± 0.76",	"0.36 ± 0.51",
  "Procrastination",	"28.60 ± 12.00", "27.70 ± 11.30",	"32.10 ± 13.80", "39.00 ± 16.00"
)

@@

# Outputting table 
tbl_descriptives |>
  setNames(c(" ", "Full sample (n = 549)", "NC (n = 452)", "MCI (n = 86)", "Dementia (n = 11)")) |>
   tt(notes = list(
    caption, 
    a = list(i = 10, j = 1, text = footnote))
  ) |>
   group_tt(i = list("**Sex**" = 2, "**Education**" = 4)) |>
   style_tt(j = 2:5, align = "c") |>
   style_tt(i = c(2, 5), align = "l") |>
   format_tt(escape = TRUE) |>
   format_tt("notes", markdown = TRUE) |>
   format_tt(j = 1, markdown = TRUE) |>
  style_tt(
     bootstrap_class = "table table-sm"
  )

```

```{r}
#| label: fig-river
#| fig-width: 15
#| fig-height: 10
#| fig-cap: "Frequency of cognitive status transitions between HRS waves."
#| fig-pos: H
#| include: true

@@

```{r}
#| label: fig-markov
#| fig-width: 12
#| fig-height: 10
#| fig-cap: "Odds ratios (95% Confidence Intervals) of transitioning from one cognitive state to another calculated using generalised logit models. *Note*. Odds ratios significantly different from 1 at a 5% significance rate are presented in blue (positive) or orange (negative),  otherwise they are presented in gray. NC = Normative cognitive function;  MCI = Mild cognitive impairment."
#| include: true

@@

```{r}
#| label: fig-age-preds
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Predicted transition probabilities by procrastination and age. *Note.* Each curve represents a different age profile, ranging from the minimum (62 years) to the maximum (97 years) of the dataset. Specific ages of interest (70, 80, and 90) are highlighted."
#| include: true
